### 2018-03-14 paper/2-background.tex
Rearrange the order of subsections in S3

The user starts by checking out the project sources, whose hashes are (for
simplicity)~1,~2~and~3, and requests to build \cmd{main.exe}, see
Fig.~\ref{fig-bazel}(a,b). By looking up the history of all previous builds of
the project\footnote{Here we ignore the issue of limited cloud storage resources
for the sake of simplicity; in practice, old entries are regularly evicted from
the storage, as further discussed in~\S\ref{sec-cloud-aspects}.}, the build
system finds that someone has already compiled these exact sources before and
their resulting files \cmd{util.o} and \cmd{main.o} had hashes 4~and~5.
Similarly, the build system finds that the hash of the resulting \cmd{main.exe}
should be 6 and downloads the actual binary from the shared cloud storage, since
it is the end build product and must therefore be materialised.

@@

\begin{figure}[t]
\begin{subfigure}[b]{0.25\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
\caption{Checkout source files}
\end{subfigure}
\begin{subfigure}[b]{0.40\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
\caption{Build \cmd{main.exe}}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
\caption{Modify \cmd{util.c} and rebuild}
\end{subfigure}
\vspace{-2mm}
\caption{A cloud build example: (a)~checkout sources, (b)~download \cmd{main.exe}
from the cloud and skip intermediate files (only their hashes are needed),
(c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which requires building
\cmd{util.o} (since nobody has compiled \cmd{util.c} before) and downloading
\cmd{main.o}. File hashes are shown inside circles.
\label{fig-bazel}}
\vspace{-2mm}
\end{figure}

@@

\begin{table}[h]
\vspace{-1mm}
\smaller
\centering
\begin{tabular}{l||l|l||l|c|c|c}
\hline
$\!$Build system$\!$& Metadata                       & Algorithm    & Dependencies & Minimal & Cutoff & Cloud$\!$\\\hline
$\!$\Make       $\!$& File modification times        & Topological  & Static       & Yes     & No     & No   $\!$\\
$\!$\Excel      $\!$& Dirty cells, calculation chain & Reordering   & Dynamic      & No      & No     & No   $\!$\\
$\!$\Shake      $\!$& Previous dependency graph      & Recursive    & Dynamic      & Yes     & Yes    & No   $\!$\\
$\!$\Bazel      $\!$& All dependency graphs, cache   & Topological  & Static       & Yes     & Yes    & Yes  $\!$\\\hline
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Summary of build system differences.\label{tab-summary}}
\vspace{-8mm}
\end{table}

>>>>

The user starts by checking out the project sources, whose hashes are (for
simplicity)~1,~2~and~3, and requests to build \cmd{main.exe}, see
Fig.~\ref{fig-bazel}(a,b). By looking up the global history of all previous
builds of the project\footnote{Here we ignore the issue of limited cloud storage
resources for the sake of simplicity; in practice, old entries are regularly
evicted from the storage, as further discussed in~\S\ref{sec-cloud-aspects}.},
the build system finds that someone has already compiled these exact sources
before and their resulting files \cmd{util.o} and \cmd{main.o} had hashes 4~and~5.
Similarly, the build system finds that the hash of the resulting \cmd{main.exe}
should be 6 and downloads the actual binary from the shared cloud storage, since
it is the end build product and must therefore be materialised.

@@

\begin{figure}[t]
\begin{subfigure}[b]{0.25\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
\vspace{-1mm}
\caption{Checkout source files}
\end{subfigure}
\begin{subfigure}[b]{0.40\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
\vspace{-1mm}
\caption{Build \cmd{main.exe}}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
\vspace{-1mm}
\caption{Modify \cmd{util.c} and rebuild}
\end{subfigure}
\vspace{-2mm}
\caption{A cloud build example: (a)~checkout sources, (b)~download \cmd{main.exe}
from the cloud and skip intermediate files (only their hashes are needed),
(c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which requires building
\cmd{util.o} (since nobody has compiled \cmd{util.c} before) and downloading
\cmd{main.o}. File hashes are shown inside circles.
\label{fig-bazel}}
\vspace{-1mm}
\end{figure}

@@

\begin{table}[h]
\vspace{-2mm}
\smaller
\centering
\begin{tabular}{l||l|l||l|c|c|c}
\hline
$\!$Build system$\!$& Metadata                       & Algorithm    & Dependencies & Minimal & Cutoff & Cloud$\!$\\\hline
$\!$\Make       $\!$& File modification times        & Topological  & Static       & Yes     & No     & No   $\!$\\
$\!$\Excel      $\!$& Dirty cells, calculation chain & Reordering   & Dynamic      & No      & No     & No   $\!$\\
$\!$\Shake      $\!$& Previous dependency graph      & Recursive    & Dynamic      & Yes     & Yes    & No   $\!$\\
$\!$\Bazel      $\!$& All dependency graphs, cache   & Topological  & Static       & Yes     & Yes    & Yes  $\!$\\\hline
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Summary of build system differences.\label{tab-summary}}
\vspace{-8mm}
\end{table}

### 2018-03-14 paper/3-compute.tex
Rearrange the order of subsections in S3

Our first main idea is the task-description abstraction:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
\emph{profunctor optics} might recognise a familiar pattern. We discuss this
in~\S\ref{sec-related-optics}.} is best introduced by an example.
Consider this \Excel spreadsheet:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
  A1: 10     B1: A1 + A2
  A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}
Here cell \cmd{A1} contains the value \cmd{10}, the cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Compute Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _   _      = Nothing
\end{minted}
\vspace{1mm}
Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given
a \hs{fetch} function and a key. It pattern-matches on the key to see if it has
a task description (in the \Excel case, a formula)
for it.  If not, it returns \hs{Nothing}, indicating that there is no formula in
that cell.  If there is a formula in the cell, it computes
the value of the formula, using \hs{fetch} to find the value of any keys on
which it depends.

@@

@@

The \hs{Compute} type and the example above might seem mysterious. Why insist on
\hs{Compute} to be polymorphic in \hs{f}? Why choose the \hs{Applicative}
constraint?

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Data.Functor.Const
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Compute Applicative k v -> k -> [k]
dependencies compute = maybe [@@] getConst . compute (\k -> Const [k])
\end{minted}
\vspace{1mm}

@@

To understand how \hs{dependencies} works, let us examine its type more closely,
expanding the type synonym \hs{Compute}:

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Data.Functor.Identity
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
execute :: Compute Monad k v -> (k -> v) -> k -> Maybe v
execute compute store = fmap runIdentity . compute (pure . store)
\end{minted}
\vspace{1mm}

\noindent
Note that \hs{execute} can accept both an applicative as well as a \emph{monadic
compute} introduced in~\S\ref{sec-compute-monad}. Below we run \hs{execute} on
the above \hs{add} example. Note that \hs{execute} returns \hs{Nothing} for
inputs.

@@

\subsection{Monadic compute}\label{sec-compute-monad}

As discussed in~\S\ref{sec-background}, some build tasks have dynamic
dependencies, which are determined by values of intermediate computations. In
our framework, such build tasks correspond to \emph{monadic compute}, i.e.
the type \hs{Compute}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
example:

\noindent
The task corresponding to the cell \cmd{B1} statically depends on \cmd{C1} and
dynamically depends on either \cmd{A1} or \cmd{A2}. We can express this using
our compute abstraction as follows\footnote{The spreadsheet example that uses
the \hs{INDIRECT} function can be expressed very similarly: simply replace the
last line of the \hs{select} function with \hs{fetch ("A" ++ show c1)}.}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
select :: Compute Monad String Integer
select fetch key | key /= "B1" = Nothing
                 | otherwise   = Just $ do
                     c1 <- fetch "C1"
                     if c1 == 1 then fetch "A1" else fetch "A2"
\end{minted}
\vspace{1mm}

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Control.Monad.Writer
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
track :: Compute Monad k v -> (k -> v) -> k -> Maybe (v, [k])
track compute store = fmap runWriter . compute (\k -> writer (store k, [k]))
\end{minted}
\vspace{1mm}

>>>>

Our first main idea is the task-description abstraction:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
\emph{profunctor optics} might recognise a familiar pattern. We discuss this
in~\S\ref{sec-related-optics}.} is best introduced by an example.
Consider this \Excel spreadsheet:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
  A1: 10     B1: A1 + A2
  A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}
Here cell \cmd{A1} contains the value \cmd{10}, the cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Task Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _   _      = Nothing
\end{minted}
\vspace{1mm}
Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given
a \hs{fetch} function and a key. It pattern-matches on the key to see if it has
a task description (in the \Excel case, a formula)
for it.  If not, it returns \hs{Nothing}, indicating that there is no formula in
that cell.  If there is a formula in the cell, it computes
the value of the formula, using \hs{fetch} to find the value of any keys on
which it depends.

@@

\subsection{Monadic tasks}\label{sec-task-monad}

As discussed in~\S\ref{sec-background}, some build tasks have dynamic
dependencies, which are determined by values of intermediate computations. In
our framework, such build tasks correspond to \emph{monadic task}, i.e.
the type \hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
example:

\noindent
The task corresponding to the cell \cmd{B1} statically depends on \cmd{C1} and
dynamically depends on either \cmd{A1} or \cmd{A2}. We can express this using
our task abstraction as follows\footnote{The spreadsheet example that uses
the \hs{INDIRECT} function can be expressed very similarly: simply replace the
last line of the \hs{select} function with \hs{fetch ("A" ++ show c1)}.}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
select :: Task Monad String Integer
select fetch key | key /= "B1" = Nothing
                 | otherwise   = Just $ do
                     c1 <- fetch "C1"
                     if c1 == 1 then fetch "A1" else fetch "A2"
\end{minted}
\vspace{1mm}

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Control.Monad.Writer
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
track :: Task Monad k v -> (k -> v) -> k -> Maybe (v, [k])
track task store = fmap runWriter . task (\k -> writer (store k, [k]))
\end{minted}
\vspace{1mm}

@@

The \hs{Task} type and the example above might seem mysterious. Why insist on
\hs{Task} to be polymorphic in \hs{f}? Why choose the \hs{Applicative}
constraint?

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Data.Functor.Const
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Task Applicative k v -> k -> [k]
dependencies task = maybe [@@] getConst . task (\k -> Const [k])
\end{minted}
\vspace{1mm}

@@

To understand how \hs{dependencies} works, let us examine its type more closely,
expanding the type synonym \hs{Task}:

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Data.Functor.Identity
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
execute :: Task Monad k v -> (k -> v) -> k -> Maybe v
execute compute store = fmap runIdentity . compute (pure . store)
\end{minted}
\vspace{1mm}

\noindent
Note that \hs{execute} can accept both an applicative as well as a \emph{monadic
task} introduced in~\S\ref{sec-task-monad}. Below we run \hs{execute} on
the above \hs{add} example. Note that \hs{execute} returns \hs{Nothing} for
inputs.

@@
